SFJ: An implementation of Semantic Featherweight Java

### 2020-04-27 conference/camera ready/paper.tex
on future work
commenting on algorithm

\begin{algorithm}[t]
    \caption{Semantic Subtyping for SFJ---\emph{generateRelation}}
    \label{alg:subtyping}
    \SetKwBlock{Begin}{begin}{}
    \DontPrintSemicolon
    \Input{$classes$, the set of classes in an SFJ program.\\
    $relation$, the mapping of types to the set of subtypes, initially Map \ref{eq:initialSubtype}. \;
    }
    \Begin{
        \Fn{generateRelation(classes: List<Class>)}{
            $unprocessed: List<Class>\ \longleftarrow []$

            \For{class in classes}{
                \If{addClass(class) = $False$}{
                    $unprocessed.add(class)$
                }
            }
            \If{$untyped \neq []$}{
                $generateRelation(unprocessed)$
            }
        }
        \;
    }
\end{algorithm}
%%
%%
\setcounter{algocf}{1}
\begin{algorithm}
    \caption{Semantic Subtyping for SFJ---\emph{addClass} and \emph{checkSuperSet}}

    \DontPrintSemicolon
    \setcounter{AlgoLine}{12}
    \SetKwBlock{Begin}{}{end}
    \Begin{
        \Fn{addClass(class: Class) $\rightarrow$ boolean}{
            \For{existing class type in relation}{
                \If{checkSuperSet(class, existingClass) = False}{
                    \KwRet $False$\;
                }
                $checkSuperSet(existingClass, class)$\;
            }
            $relation[class].add(class)$ \;
            $relation[Universe].add(class)$ \;
            \KwRet $True$ \;
        }
        \;
        \Fn{checkSuperSet(class: Class, other: Class) $\rightarrow$ boolean}{
            $flag \longleftarrow True$ \;
            \For{field in class}{
                \If{field contains type not in relation}{
                    \KwRet $False$
                }
                \uIf{other does not contain field}{
                    $flag \longleftarrow False$
                }
                \Else{
                    \If{other.field.types does not fully contain field.types}{
                        $flag \longleftarrow False$
                    }
                }
            }

@@

Finally, the subtyping relation generated by \autoref{alg:subtyping} is used in the type system for the SFJ language via a subsumption typing rule:
 $$
 \frac{\Gamma\vdash e:\alpha_1 \qquad \alpha_1 \leq \alpha_2}{\Gamma \vdash e:\alpha_2}
 $$
We read this typing rule as follows:
if an expression $e$ is of type $\alpha_1$ under a typing context $\Gamma$ (details of a typing context are irrelevant here) and type $\alpha$ is a subtype $\leq$ of $\alpha_2$, then expression $e$ can be typed with $\alpha_2$.

@@

Let us illustrate the semantic subtyping algorithm on our \emph{Polygons} given in \autoref{polygons}.
\autoref{alg:subtyping} generates the subtyping relation given in \autoref{eq:sugtypePolygons}, together with the subtyping relation for basic types, omitted here and defined in \autoref{eq:initialSubtype}. Notice that the mapping for $Universe$ is extended with the new types for polygons.
\begin{equation}
    \label{eq:sugtypePolygons}
    \begin{array}{llllll}
         & Polygon  & = \left\{Polygon, Triangle, Square, Rhombus \right\} &  & Triangle & = \left\{Triangle\right\}           \\
         & Square   & = \left\{Square\right\}                              &  & Rhombus  & = \left\{Rhombus\right\}            \\
         & Universe & = \{Double, Float, Long, Int, Short, Byte                                                              &  & Diagonal & = \left\{Diagonal\right\}          \\
         &          & \quad Boolean, Void, Polygon, Square                                                                     \\
         &          & \quad Square, Rhombus, Diagonal\}
    \end{array}
\end{equation}

Recall the method \emph{diagonal} in class \emph{Diagonal}, with signature
$$
\mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)
$$
The result of the set operation on its parameter type gives the following set of polygons:
$$
Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
$$
In order to define the \textbf{not} $Triangle$ type we need the $Universe$ type so that we can define it as $Universe\ \mathbf{\setminus}\ Triangle$. Then, the \textbf{and} connective is the intersection of sets of $Polygon$ with \textbf{not} $Triangle$.

@@

\section{Conclusion}
\label{sec:conclusion}

In this paper we presented the design and implementation of SFJ---Semantic Featherweight Java, an extension of Featherweight Java featuring boolean types, semantic subtyping, and integrating both nominal and structural subtyping. Due to the expressivity of semantic subtyping, in SFJ we are able to restore standard Java constructs and features for example, lists and overloading meathods, which were not present in FJ, thus making SFJ a more complete language.
We presented \autoref{alg:types} on validity of type definitions and
\autoref{alg:subtyping} on semantic subtyping, which finds all nominal and structural subtypes for all types in an SFJ program. We also described typing of terms in SFJ, which follows that of Java and builds upon relevant work \cite{Dardha2013,Dardha2017}.

@@

>>>>

\begin{algorithm}[t]
    \caption{Semantic Subtyping for SFJ---\emph{generateRelation}}
    \label{alg:subtyping}
    \SetKwBlock{Begin}{begin}{}
    \DontPrintSemicolon
    \SetNoFillComment
    \Input{$classes$, the set of classes in an SFJ program.\\
    $relation$, the mapping of types to the set of subtypes, initially Map \ref{eq:initialSubtype}. \;
    }
    \Begin{
        \Fn{generateRelation(classes: List<Class>)}{
            $unprocessed: List<Class>\ \longleftarrow []$

            \For{class in classes}{
                \If{addClass(class) = $False$}{
                    $unprocessed.add(class)$
                }
            }
            \If{$untyped \neq []$}{
                $generateRelation(unprocessed)$
            }
        }
        \;
    }
    \tcc{algorithm continued on next page...}
\end{algorithm}
%%
%%
\setcounter{algocf}{1}
\begin{algorithm}
    \caption{Semantic Subtyping for SFJ---\emph{addClass} and \emph{checkSuperSet}}

    \DontPrintSemicolon
    \SetNoFillComment
    \tcc{...algorithm continued from previous page}

    \setcounter{AlgoLine}{12}
    \SetKwBlock{Begin}{}{end}
    \Begin{
        \Fn{addClass(class: Class) $\rightarrow$ boolean}{
            \For{existing class type in relation}{
                \If{checkSuperSet(class, existingClass) = False}{
                    \KwRet $False$\;
                }
                $checkSuperSet(existingClass, class)$\;
            }
            $relation[class].add(class)$ \;
            $relation[Universe].add(class)$ \;
            \KwRet $True$ \;
        }
        \;
        \Fn{checkSuperSet(class: Class, other: Class) $\rightarrow$ boolean}{
            $flag \longleftarrow True$ \;
            \For{field in class}{
                \If{field contains type not in relation}{
                    \KwRet $False$
                }
                \uIf{other does not contain field}{
                    $flag \longleftarrow False$
                }
                \Else{
                    \If{other.field.types does not fully contain field.types}{
                        $flag \longleftarrow False$
                    }
                }
            }

@@

Finally, the subtyping relation generated by \autoref{alg:subtyping} is used in the type system for the SFJ language via a subsumption typing rule:
 $$
 \frac{\Gamma\vdash e:\alpha_1 \qquad \alpha_1 \leq \alpha_2}{\Gamma \vdash e:\alpha_2}
 $$

We read this typing rule as follows:
if an expression $e$ is of type $\alpha_1$ under a typing context $\Gamma$ (details of a typing context are irrelevant here) and type $\alpha$ is a subtype $\leq$ of $\alpha_2$, then expression $e$ can be typed with $\alpha_2$.

@@

Let us illustrate the semantic subtyping algorithm on our \emph{Polygons} given in \autoref{polygons}.
\autoref{alg:subtyping} generates the subtyping relation given in \autoref{eq:sugtypePolygons}, together with the subtyping relation for basic types, omitted here and defined in \autoref{eq:initialSubtype}. Notice that the mapping for $Universe$ is extended with the new types for polygons.
\begin{equation}
    \label{eq:sugtypePolygons}
    \begin{array}{llllll}
         & Polygon  & = \left\{Polygon, Triangle, Square, Rhombus \right\} &  & Triangle & = \left\{Triangle\right\}           \\
         & Square   & = \left\{Square\right\}                              &  & Rhombus  & = \left\{Rhombus\right\}            \\
         & Universe & = \{Double, Float, Long, Int, Short, Byte                                                              &  & Diagonal & = \left\{Diagonal\right\}          \\
         &          & \quad Boolean, Void, Polygon, Square                                                                     \\
         &          & \quad Square, Rhombus, Diagonal\}
    \end{array}
\end{equation}
%
Recall the method \emph{diagonal} in class \emph{Diagonal}, with signature
$$
\mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)
$$
The result of the set operation on its parameter type gives the following set of polygons:
$$
Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
$$
In order to define the \textbf{not} $Triangle$ type we need the $Universe$ type so that we can define it as $Universe\ \mathbf{\setminus}\ Triangle$. Then, the \textbf{and} connective is the intersection of sets of $Polygon$ with \textbf{not} $Triangle$.

@@

\section{Related Work and Conclusion}
\label{sec:conclusion}

@@

In this paper we presented the design and implementation of SFJ---Semantic Featherweight Java, an extension of Featherweight Java featuring boolean types, semantic subtyping, and integrating both nominal and structural subtyping. Due to the expressivity of semantic subtyping, in SFJ we are able to restore standard Java constructs and features for example, lists and overloading meathods, which were not present in FJ, thus making SFJ a more complete language.
We presented \autoref{alg:types} on validity of type definitions and
\autoref{alg:subtyping} on semantic subtyping, which finds all nominal and structural subtypes for all types in an SFJ program. We also described typing of terms in SFJ, which follows that of Java and builds upon relevant work \cite{Dardha2013,Dardha2017}.
As future work, we aim to finalise the code generation phase, which is sketched in \autoref{sec:code_generation}.
